Type Classes for Mathematics in Type Theory

### 2010-06-28 jar.tex
Fix some citations; other fixes.

Effectively keeping track of, resolving, and combining proofs of equivalence-ness and propriety when the user attempts to substitute a given (sub)term using a given equality, is known as ``setoid rewriting'', and requires nontrivial infrastructure and support from the system. The Coq implementation of these mechanisms was largely rewritten in 200? by Matthieu Sozeau~\cite{Setoid-rewrite} to make it more flexible and to replace the old special-purpose setoid/morphism registration command with a clean type class based interface.

@@

In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{C-corn} and Ssreflect\ref{Packed}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.

@@

Manifest fields~\ref{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.

@@

Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then ``lifted'' to the abstract interface. For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations. For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using involutivity of \lstinline|naturals_to_semiring|.\marginpar{Word: Involutive?}

>>>>

Effectively keeping track of, resolving, and combining proofs of equivalence-ness and propriety when the user attempts to substitute a given (sub)term using a given equality, is known as ``setoid rewriting'', and requires nontrivial infrastructure and support from the system. The Coq implementation of these mechanisms was largely rewritten by Matthieu Sozeau~\cite{Setoid-rewrite} to make it more flexible and to replace the old special-purpose setoid/morphism registration command with a clean type class based interface.

@@

In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\cite{C-corn} and Ssreflect\cite{Packed}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.

@@

Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.

@@

Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then ``lifted'' to the abstract interface. For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations. For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using the fact that \lstinline|naturals_to_semiring| is an involution.
